---
title: "Time Series Example, Reference: Data Science and Big Data Analytics"
output: html_notebook
---



```{r}
# read in gasoline production time series
# monthly gas production expressed in millions of barrels
library(forecast)
gas_prod_input <- as.data.frame( read.csv("gas_prod.csv") )

# create a time series object
gas_prod <- ts(gas_prod_input[,2])

#examine the time series
plot(gas_prod, xlab = "Time (months)",
     ylab = "Gasoline production (millions of barrels)")
```

```{r}
# check for conditions of a stationary time series
plot(diff(gas_prod))
abline(a=0, b=0)

```

The differenced time series has a constant mean around 0 with fairly constant variance over time. This is a stationary time series. We can now apply the AR models. Let us examine, ACH, and PACH
```{r}
# examine ACF and PACF of differenced series
acf(diff(gas_prod), xaxp = c(0, 48, 4), lag.max=48, main="")
pacf(diff(gas_prod), xaxp = c(0, 48, 4), lag.max=48, main="")
```

What do you notice from the ACH and PACH plots?

Let us apply a seasonal AR(1) model with period = 12. We address the seasonal component before addressing the non-seasonal components
```{r}
# fit a (0,1,0)x(1,0,0)12 ARIMA model
arima_1 <- arima (gas_prod,
                  order=c(0,1,0), # this specifies the differencing d = 1
                  seasonal = list(order=c(1,0,0),period=12))
arima_1
# it may be necessary to calculate AICc and BIC 
# http://stats.stackexchange.com/questions/76761/extract-bic-and-aicc-from-arima-object
AIC(arima_1,k = log(length(gas_prod)))   #BIC

```

What can you conclude from the coefficient value?

Let us examine the residuals from the arima_1 model
```{r}
# examine ACF and PACF of the (0,1,0)x(1,0,0)12 residuals
acf(arima_1$residuals, xaxp = c(0, 48, 4), lag.max=48, main="")
pacf(arima_1$residuals, xaxp = c(0, 48, 4), lag.max=48, main="")

```

What do your observe? Note the lack of correlation at lags  12, 24, 36, and 48. But, there is correlation at lag 1. Also, the PACH shows slow decay starting at lag 1. This suggests a MA(1) model for the nonseasonal component. In other words, let us consider (0,1,1) x (1,0,0)12 ARIMA model to the original time series
```{r}
# fit a (0,1,1)x(1,0,0)12 ARIMA model
arima_2 <- arima (gas_prod,
                  order=c(0,1,1),
                  seasonal = list(order=c(1,0,0),period=12))
arima_2

# it may be necessary to calculate AICc and BIC 
# http://stats.stackexchange.com/questions/76761/extract-bic-and-aicc-from-arima-object
AIC(arima_2,k = log(length(gas_prod)))   #BIC

```
```{r}
# examine ACF and PACF of the (0,1,1)x(1,0,0)12 residuals
acf(arima_2$residuals, xaxp = c(0, 48, 4), lag.max=48, main="")
pacf(arima_2$residuals, xaxp = c(0, 48,4), lag.max=48, main="")

```

Now, let us use the arima_2 model to forecast
```{r}
# Forecasting

#predict the next 12 months
arima_2.predict <- predict(arima_2,n.ahead=12)
matrix(c(arima_2.predict$pred-1.96*arima_2.predict$se,
         arima_2.predict$pred,
         arima_2.predict$pred+1.96*arima_2.predict$se), 12,3,
       dimnames=list( c(241:252) ,c("LB","Pred","UB")) )

plot(gas_prod, xlim=c(145,252),
     xlab = "Time (months)",
     ylab = "Gasoline production (millions of barrels)",
     ylim=c(360,440))
lines(arima_2.predict$pred)
lines(arima_2.predict$pred+1.96*arima_2.predict$se, col=4, lty=2)
lines(arima_2.predict$pred-1.96*arima_2.predict$se, col=4, lty=2)
```

