---
title: "Time Series Forecasting"
output: html_notebook
---
References: 
Data Mining for Business Analytics, Shmueli, et al.
EMC 
Code and references have been derived from the above books.

```{r}
#install.packages("forecast")

```

```{r}
Amtrak.data <- read.csv("Amtrak.csv")
Amtrak.data
```
# create time series object using ts()
# ts() takes three arguments: start, end, and freq. 
# with monthly data, the frequency of periods per season is 12 (per year). 
# arguments start and end are (season number, period number) pairs. 
# here start is Jan 1991: start = c(1991, 1); end is Mar 2004: end = c(2004, 3).
```{r}
ridership.ts <- ts(Amtrak.data$Ridership, 
    start = c(1991, 1), end = c(2004, 3), freq = 12)
plot(ridership.ts, xlab = "Time", ylab = "Ridership (in 000s)", ylim = c(1300, 2300))
```
# create short time series
# use window() to create a new, shorter time series of ridership.ts
# for the new three-year series, start time is Jan 1997 and end time is Dec 1999
```{r}
ridership.ts.3yrs <- window(ridership.ts, start = c(1997, 1), end = c(1999, 12))
```
# fit a linear regression model to the time series
```{r}
library(forecast)
ridership.lm <- tslm(ridership.ts ~ trend + I(trend^2))
```
# shorter and longer time series

```{r}
par(mfrow = c(2, 1))
plot(ridership.ts.3yrs, xlab = "Time", ylab = "Ridership (in 000s)", 
                        ylim = c(1300, 2300))
plot(ridership.ts, xlab = "Time", ylab = "Ridership (in 000s)", ylim = c(1300, 2300))
# overlay the fitted values of the linear model
lines(ridership.lm$fitted, lwd = 1)

```
Now we do Naive forecasts
First split the data into training and validation
```{r}
nValid <- 36
nTrain <- length(ridership.ts) - nValid
# partition the data
train.ts <- window(ridership.ts, start = c(1991, 1), end = c(1991, nTrain))
valid.ts <- window(ridership.ts, start = c(1991, nTrain + 1),
                   end = c(1991, nTrain + nValid))
 
```

```{r}
#  generate the naive and seasonal naive forecasts
naive.pred <- forecast::naive(train.ts, h = nValid)
snaive.pred <- forecast::snaive(train.ts, h = nValid)
```

```{r}
# plot forecasts and actuals in the training and validation sets
plot(train.ts, ylim = c(1300, 2600),  ylab = "Ridership", xlab = "Time", bty = "l", 
    xaxt = "n", xlim = c(1991,2006.25), main = "") 
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(naive.pred$mean, lwd = 2, col = "blue", lty = 1)
lines(snaive.pred$mean, lwd = 2, col = "red", lty = 1)
lines(valid.ts, col = "grey20", lty = 3)
lines(c(2004.25 - 3, 2004.25 - 3), c(0, 3500)) 
lines(c(2004.25, 2004.25), c(0, 3500))
text(1996.25, 2500, "Training")
text(2002.75, 2500, "Validation")
text(2005.25, 2500, "Future")
arrows(2004 - 3, 2450, 1991.25, 2450, code = 3, length = 0.1, lwd = 1,angle = 30)
arrows(2004.5 - 3, 2450, 2004, 2450, code = 3, length = 0.1, lwd = 1,angle = 30)
arrows(2004.5, 2450, 2006, 2450, code = 3, length = 0.1, lwd = 1, angle = 30)
```

```{r}
forecast::accuracy(naive.pred, valid.ts)
forecast::accuracy(snaive.pred, valid.ts)
```

Now, let us model the time series using regression models
Linear Trend
```{r}
library(forecast)

#Amtrak.data <- read.csv("Amtrak.csv")

# create time series

ridership.ts <- ts(Amtrak.data$Ridership, start = c(1991,1),
                   end = c(2004,3), freq = 12)

# produce linear trend model
ridership.lm <- tslm(ridership.ts ~ trend)

# plot the series
plot(ridership.ts, xlab = "Time", ylab = "Ridership", ylim = c(1300,2300),
     bty = "l")
lines(ridership.lm$fitted, lwd = 2)
```

Now, run the model using validation and training sets
```{r}
# fit linear trend model to training set and create forecasts
train.lm <- tslm(train.ts ~ trend)
train.lm.pred <- forecast(train.lm, h = nValid, level = 0)

par(mfrow = c(2, 1))
plot(train.lm.pred, ylim = c(1300, 2600),  ylab = "Ridership", xlab = "Time", 
     bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "", flty = 2)
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(train.lm.pred$fitted, lwd = 2, col = "blue")
lines(valid.ts)
plot(train.lm.pred$residuals, ylim = c(-420, 500),  ylab = "Forecast Errors", 
    xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "")
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(valid.ts - train.lm.pred$mean, lwd = 1)
```
Summmary
```{r}
summary(train.lm)
```

Now, let's fit an exponential trend
```{r}
# fit exponential trend using tslm() with argument lambda = 0 
train.lm.expo.trend <- tslm(train.ts ~ trend, lambda = 0)
train.lm.expo.trend.pred <- forecast(train.lm.expo.trend, h = nValid, level = 0)

# fit linear trend using tslm() with argument lambda = 1 (no transform of y)
train.lm.linear.trend <- tslm(train.ts ~ trend, lambda = 1)
train.lm.linear.trend.pred <- forecast(train.lm.linear.trend, h = nValid, level = 0)

plot(train.lm.expo.trend.pred, ylim = c(1300, 2600),  ylab = "Ridership", 
 xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "", flty = 2)
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(train.lm.expo.trend.pred$fitted, lwd = 2, col = "blue")  # Added in 6-5
lines(train.lm.linear.trend.pred$fitted, lwd = 2, col = "black", lty = 3)
lines(train.lm.linear.trend.pred$mean, lwd = 2, col = "black", lty = 3)
lines(valid.ts)
```

Now, fit a quadratic trend
```{r}
#fit quadratic trend using function I(), which treats an object "as is".
train.lm.poly.trend <- tslm(train.ts ~ trend + I(trend^2))
summary(train.lm.poly.trend)
train.lm.poly.trend.pred <- forecast(train.lm.poly.trend, h = nValid, level = 0)

par(mfrow = c(2,1))
plot(train.lm.poly.trend.pred, ylim = c(1300, 2600),  ylab = "Ridership", 
 xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "", flty = 2)
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1))) 
lines(train.lm.poly.trend$fitted, lwd = 2)
lines(valid.ts)

plot(train.lm.poly.trend$residuals, ylim = c(-400, 550),  ylab = "Forecast Errors", 
 xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "")
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(valid.ts - train.lm.poly.trend.pred$mean, lwd = 1)
```

Now, let us add seasonality
Our new file has a column called Season
```{r}
AmtrakTS.data <- read.csv("AmtrakTS.csv")
AmtrakTS.data

```
First split the data into training and validation
```{r}
ridership.ts <- ts(AmtrakTS.data$Ridership, 
    start = c(1991, 1), end = c(2004, 3), freq = 12)
nValid <- 36
nTrain <- length(ridership.ts) - nValid
# partition the data
train.ts <- window(ridership.ts, start = c(1991, 1), end = c(1991, nTrain))
valid.ts <- window(ridership.ts, start = c(1991, nTrain + 1),
                   end = c(1991, nTrain + nValid))
```

# include season as a predictor in tslm(). Here it creates 11 dummies
```{r}
train.lm.season <- tslm(train.ts ~ season)
summary(train.lm.season)
```

Now add trend to the season data
```{r}
train.lm.trend.season <- tslm(train.ts ~ trend + I(trend^2) + season)
summary(train.lm.trend.season)
train.lm.trend.season.pred <- forecast(train.lm.trend.season, h = nValid, level = 0)

par(mfrow = c(2,1))
plot(train.lm.trend.season.pred, ylim = c(1300, 2600),  ylab = "Ridership", 
 xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "", flty = 2)
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1))) 
lines(train.lm.trend.season$fitted, lwd = 2)
lines(valid.ts)

plot(train.lm.trend.season$residuals, ylim = c(-400, 550),  ylab = "Forecast Errors", 
 xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "")
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(valid.ts - train.lm.trend.season.pred$mean, lwd = 1, col="red")
```
Now let's calculate the autocorrelation
```{r}
ridership.24.ts <- window(train.ts, start = c(1991, 1), end = c(1991, 24)) 
Acf(ridership.24.ts, lag.max = 12, main = "")
# We can also look at the autocorrelation of the residuals
# For example, we calculate the autocorrelation of the residuals after fitting for season and trend
Acf(train.lm.trend.season$residuals, lag.max = 12, main="")
#Let us also display partial autocorrelation
Pacf(train.lm.trend.season$residuals, lag.max = 12, main="")
```
```{r}
# fit linear regression with quadratic trend and seasonality to Ridership 
train.lm.trend.season <- tslm(train.ts ~ trend + I(trend^2) + season) 
# fit AR(1) model to training residuals 
# use Arima() in the forecast package to fit an ARIMA model # (that includes AR models); order = c(1,0,0) gives an AR(1). 
train.res.arima <- Arima(train.lm.trend.season$residuals, order = c(1,0,0)) 
valid.res.arima.pred <- forecast(train.res.arima, h = 1) 
#Output > 
summary(train.res.arima)
```

Thus, our AR(1) model is 
Et = 0.3728 + 0.5998 * E(t-1) + epsilon

To now forecast residual for April 2001, we plug in the residual from the most recent residual from March 2001 (12.108) into the AR(1) model, this gives

0.3728 + 0.5998 * (12.108 - 0.3728)**

The answer is 7.411. We can get this directly using the forecast() function

** Note: The intercept in the Coefficients table resulting from function Arima() is not exactly an intercept—it is the estimated mean of the series. Hence, to get a forecast, we must subtract this coefficient from our value. In this case, we have Ft + 1 = intercept + slope (yt − intercept).
```{r}
valid.res.arima.pred
```

The positive value indicates that the regression will produce a forecasted value for April 2001 that is too low, and that we should add 7411 riders to the forecasted value. In this particular example, the regression model (with quadratic trend and seasonality) produced a forecast of 2,004,271 riders, and the improved two-stage model [regression + AR(1) correction] corrected it by increasing it to 2,011,906 riders. The actual value for April 2001 turned out to be 2,023,792 riders—much closer to the improved forecast.

From the plot of the actual vs. forecasted residual series (below), we can see that the AR(1) model fits the residual series quite well. Note, however, that the plot is based on the training data (until March 2001). To evaluate predictive performance of the two-level model [regression + AR(1)], we would have to examine performance (e.g., via MAPE or RMSE metrics) on the validation data, in a fashion similar to the calculation that we performed for April 2001.
```{r}
plot(train.lm.trend.season$residuals, ylim = c(-250, 250), ylab = "Residuals", 
     xlab = "Time", bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "")
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(train.res.arima$fitted, lwd = 2, col = "blue")
```

Now, let us consider data-driven methods. Specifically, smoothing methods.

Let us look at moving averages

How should w be chosen for a seasonal series?

```{r}
Amtrak.data <- read.csv("Amtrak.csv")
Amtrak.data
ridership.ts <- ts(Amtrak.data$Ridership, 
    start = c(1991, 1), end = c(2004, 3), freq = 12)
```
```{r}
library(zoo)
# centered moving average with window order = 12
ma.centered <- ma(ridership.ts, order = 12)

# generate a plot 
plot(ridership.ts, ylim = c(1300, 2200),  ylab = "Ridership", 
    xlab = "Time", bty = "l", xaxt = "n", 
    xlim = c(1991,2004.25), main = "")
axis(1, at = seq(1991, 2004.25, 1), labels = format(seq(1991, 2004.25, 1)))
lines(ma.centered, lwd = 2)
legend(1994,2200, c("Ridership","Centered Moving Average"), 
    lty=c(1,1), lwd=c(1,2), bty = "n")  
```

Now, let's plot a trailing moving average

```{r}
# centered moving average with window order = 12
ma.centered <- ma(ridership.ts, order = 12)

# trailing moving average with window k = 12
# in rollmean(), use argument align = right to calculate a trailing moving average.
ma.trailing <- rollmean(ridership.ts, k = 12, align = "right")

# generate a plot 
plot(ridership.ts, ylim = c(1300, 2200),  ylab = "Ridership", 
    xlab = "Time", bty = "l", xaxt = "n", 
    xlim = c(1991,2004.25), main = "")
axis(1, at = seq(1991, 2004.25, 1), labels = format(seq(1991, 2004.25, 1)))
lines(ma.centered, lwd = 2)
lines(ma.trailing, lwd = 2, lty = 2)
legend(1994,2200, c("Ridership","Centered Moving Average", "Trailing Moving Average"), 
    lty=c(1,1,2), lwd=c(1,2,2), bty = "n")  

```

Next, we illustrate a 12-month moving average forecaster for the Amtrak ridership. We partition the Amtrak ridership time series, leaving the last 36 months as the validation period. Applying a moving average forecaster with window w = 12. Note that for the first 12 records of the training period, there is no forecast (because there are less than 12 past values to average). Also, note that the forecasts for all months in the validation period are identical (1938.481) because the method assumes that information is known only until March 2001.
```{r}
# partition the data
nValid <- 36
nTrain <- length(ridership.ts) - nValid
train.ts <- window(ridership.ts, start = c(1991, 1), end = c(1991, nTrain))
valid.ts <- window(ridership.ts, start = c(1991, nTrain + 1), 
    end = c(1991, nTrain + nValid))

# moving average on training
ma.trailing <- rollmean(train.ts, k = 12, align = "right")

# obtain the last moving average in the training period
last.ma <- tail(ma.trailing, 1)

# create forecast based on last MA
ma.trailing.pred <- ts(rep(last.ma, nValid), start = c(1991, nTrain + 1), 
    end = c(1991, nTrain + nValid), freq = 12)
    
# plot the series
plot(train.ts, ylim = c(1300, 2600),  ylab = "Ridership", xlab = "Time", bty = "l", 
    xaxt = "n", xlim = c(1991,2006.25), main = "")
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(ma.trailing, lwd = 2, col = "blue") 
lines(ma.trailing.pred, lwd = 2, col = "blue", lty = 2)
lines(valid.ts)
```

Consider the regression model shown in Figure 17.6 in Chapter 17, which yields residuals devoid of seasonality and trend (see bottom chart). We can apply a moving average forecaster to that series of residuals (also called forecast errors), thereby creating a forecast for the next forecast error. 
```{r}
AmtrakTS.data <- read.csv("AmtrakTS.csv")
AmtrakTS.data
ridership.ts <- ts(AmtrakTS.data$Ridership, 
    start = c(1991, 1), end = c(2004, 3), freq = 12)
nValid <- 36
nTrain <- length(ridership.ts) - nValid
# partition the data
train.ts <- window(ridership.ts, start = c(1991, 1), end = c(1991, nTrain))
valid.ts <- window(ridership.ts, start = c(1991, nTrain + 1),
                   end = c(1991, nTrain + nValid))
# fit regression model with trend and seasonality
train.lm.trend.season <- tslm(train.ts ~ trend + I(trend^2) + season)

# create single-point forecast
train.lm.trend.season.pred <- forecast(train.lm.trend.season, h = 1, level = 0)

# apply MA to residuals
ma.trailing <- rollmean(train.lm.trend.season$residuals, k = 12, align = "right")
last.ma <- tail(ma.trailing, 1)
train.lm.trend.season.pred
last.ma
```

Now, let's apply exponential smoothing to the residuals from our previous model. The residuals are assumed to have no trend or seasonality
```{r}
# get residuals
residuals.ts <- train.lm.trend.season$residuals

# run simple exponential smoothing
# use ets() with model = "ANN" (additive error (A), no trend (N), no seasonality (N)) 
# and alpha = 0.2 to fit simple exponential smoothing.
ses <- ets(residuals.ts, model = "ANN", alpha = 0.2)
ses.pred <- forecast(ses, h = nValid, level = 0)

plot(ses.pred, ylim = c(-250, 300),  ylab = "Ridership", xlab = "Time", 
     bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "", flty = 2)
lines(train.lm.trend.season.pred$fitted, lwd = 2, col = "blue")
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(ses.pred$fitted, lwd = 2, col = "blue")
lines(valid.ts)
```

Now, let us fit the Holt-Winters exponential model
```{r}
# run Holt-Winters exponential smoothing
# use ets() with option model = "MAA" to fit Holt-Winter's exponential smoothing 
# with multiplicative error, additive trend, and additive seasonality. 
hwin <- ets(train.ts, model = "MAA")

# create predictions
hwin.pred <- forecast(hwin, h = nValid, level = 0)

# plot the series
plot(hwin.pred, ylim = c(1300, 2600),  ylab = "Ridership", xlab = "Time", 
    bty = "l", xaxt = "n", xlim = c(1991,2006.25), main = "", flty = 2)
axis(1, at = seq(1991, 2006, 1), labels = format(seq(1991, 2006, 1)))
lines(hwin.pred$fitted, lwd = 2, col = "blue")
lines(valid.ts)
```
```{r}
hwin
```

